A Type System for Safe and Scalable Parallelism using Logical Regions

### 2012-07-11 doc/popl2013/example.tex
changes to intro and code example sections

Listing~\ref{lst:circuit_ex} shows a circuit simulation written (with
one exception) in the core Legion language (see
Section~\ref{sec:legioncore}).  The input is a graph of circuit
elements and connecting wires.  To perform parallel
simulation the graph is partitioned into pieces, after which the
simulation is run for a number of time steps.  Each time step has
three phases: calculate each wire's current, distribute charge to
the nodes to which each wire is connected, and update each node's voltage.

The top-level function is {\tt simulate\_circuit} (line 15), which is
parameterized on regions {\tt rl}, {\tt rw}, and {\tt rn} (line 15).
The wires and nodes live in regions {\tt rw} and {\tt rn},
respectively.  The {\tt rl} region holds lists of the nodes and wires.
When {\tt simulate\_circuit} is invoked it must have read and
write privileges for the region parameters
(line 17); the type system enforces that
{\tt simulate\_circuit} only accesses these regions and their
subregions.

Partitioning of the graph is done using {\em colorings} built in the
{\tt color\_circuit} function (line 18-19).  A coloring maps region
elements to colors, with each color corresponding to a new subregion
to be created.  Note that it is impossible to pick a single static
partitioning scheme that would do well for all graphs.  Dynamic
partitioning using colorings provides the programmer the flexibility
to make partitioning decisions based on input data.

@@

After creating the colorings, the application partitions the region
into {\em subregions} according to the colorings (lines 21-25).  For
simplicity, our example only partitions the circuit into two pieces.
Lines 21 and 23 create two different partitions each with two different
subregions of the {\tt rn} region.  The first partition uses a
coloring and is therefore {\em disjoint}; the second partition uses a
multicoloring and is {\em aliased}---the subregions may not be
disjoint. While aliased partitions based on multicolorings are not a
part of core Legion described in Section~\ref{sec:legioncore}, they
are functionally equivalent to creating single partitions for each
color in the multicoloring.
Disjoint partitions introduce region disjointness constraints into the static
environment indicating that all the subregions in the partition are
disjoint (e.g. {\tt rn0}$*${\tt rn1} on line 21 says {\tt rn0} is
disjoint from {\tt rn1}).  Both disjoint and aliased partitions
introduce constraints into the environment indicating that all created
regions are logical subregions of the partitioned parent region
(e.g. {\tt rn0}$\leq${\tt rn}).

The circuit simulation uses an {\em allocate-then-partition} style of
computation, where a large data structure is first allocated and then
partitioned.  Legion also supports {\em partition-then-allocate},
where empty regions are first partitioned and later populated with
data.  Both are useful; e.g., Sequoia \cite{Fatahalian06} and DPJ
\cite{Bocchino09} support static forms of allocate-then-partition and
partition-then-allocate, respectively.  Legion supports both
approaches dynamically.

@@

The {\tt execute\_time\_steps} function (lines 37-48)  
also illustrates the importance of using different partitions to give
multiple views onto the same logical region.  Both {\tt
  calc\_new\_currents} and {\tt distribute\_charge} 
use the owned and ghost regions of a piece, which are from different partitions. In
{\tt calc\_new\_currents} these regions only need read
privileges, allowing both instances of {\tt calc\_new\_currents} to be
run in parallel.  In {\tt distribute\_charge} the
privilege is for a reduction which can also be done in parallel
because of the atomic and commutative nature of reductions.  No
single partition of the nodes
describes these data sharing patterns.

>>>>

Listing~\ref{lst:circuit_ex} shows a circuit simulation written (with
one exception) in the core Legion language (see
Section~\ref{sec:legioncore}).  The input is a graph of circuit
elements and connecting wires.  To perform parallel
simulation the graph is partitioned into pieces, after which the
simulation is run for a number of time steps.  Each time step has
three phases: calculation of each wire's current, distribution of charge to
the nodes to which each wire is connected, and updating of each node's voltage.

The top-level function is {\tt simulate\_circuit} (line 15), which is
parameterized on regions {\tt rl}, {\tt rw}, and {\tt rn}.
The wires and nodes are contained in regions {\tt rw} and {\tt rn},
respectively.  The {\tt rl} region holds lists of pointers to the nodes and wires.
When {\tt simulate\_circuit} is invoked it must have both read and
write privileges for the regions {\tt rl}, {\tt rw}, and {\tt rn}
(line 17); the type system enforces that
{\tt simulate\_circuit} only accesses these regions and their
subregions.

Partitioning of the graph is done using {\em colorings} built in the
{\tt color\_circuit} function (line 18-19).  A coloring maps region
elements to ``colors'' (i.e. integers), with each color corresponding
 to a new subregion
to be created.  Note that it is impossible to pick a single static
partitioning scheme that would do well for all graphs.  Dynamic
partitioning using colorings provides the programmer the flexibility
to make partitioning decisions based on input data.

@@

After creating the colorings, the application partitions the node and wire
regions
into {\em subregions} according to the colorings (lines 21-25).  For
simplicity, our example only partitions the circuit into two pieces.
Line 21 uses {\tt pc.1} to partition {\tt rn} into {\tt rn0}, and {\tt rn1}.
Subregion {\tt rn0} will contain all locations with a color of 1 in {\tt pc.1},
while {\tt rn1} will contain all locations with a color of 2.  This
partitioning uses a coloring and results in {\em disjoint} subpartitions.
Constraints are introduced into the static environment describing both the
disjointness of the subregions (i.e. {\tt rn0 $*$ rn1}) and the inclusion of
each subregion in the original region (e.g. {\tt rn0 $\leq$ rn}).  Line
23 is similar, but uses a multicoloring, so the subregions may be {\em aliased},
and only the inclusion constraints are introduced into the static environment.
Although multicolorings are not part of the Core Legion described in 
Section~\ref{sec:legioncore}, they can be easily implemented using a separate 
coloring for each color in the multicoloring.

The circuit simulation uses an {\em allocate-then-partition} style of
computation, where a large data structure is first allocated and then
partitioned.  Legion also supports a {\em partition-then-allocate} style,
where empty regions are first partitioned and later populated with
data.  Both are useful; e.g., Sequoia \cite{Fatahalian06} and DPJ
\cite{Bocchino09} support static forms of allocate-then-partition and
partition-then-allocate, respectively.  Legion supports both
approaches dynamically.

@@

The {\tt execute\_time\_steps} function (lines 37-48)  
also illustrates the importance of using different partitions to give
multiple views onto the same logical region.  Both {\tt
  calc\_new\_currents} and {\tt distribute\_charge} 
use the owned and ghost regions of a piece, which are from different partitions. In
{\tt calc\_new\_currents} these regions only need read
privileges, allowing both instances of {\tt calc\_new\_currents} to be
run in parallel.  In {\tt distribute\_charge} the
privilege is for a reduction which can also be done in parallel
because of the atomic and commutative nature of reductions.  Finally,
the {\tt update\_voltage} function modifies only the owned regions, permitting
each piece to be updated in parallel.  No
single partition of the nodes
describes these data sharing patterns.

This function also demonstrates the benefit of being able to dynamically
(re-)discover disjointness.  Safely running both instances of 
{\tt calc\_new\_currents} in parallel depends on knowing that {\tt rw0} and
{\tt rw1} in {\tt execute\_time\_steps} are disjoint.  Similarly, parallel
execution of the instances of {\tt update\_voltage} requires knowing that
{\tt rn0} is disjoint from {\tt rn1}.  In both cases, this knowledge was
statically available in {\tt simulate\_circuit} and could have been captured
in a region relationship, but it would have resulted in much more complicated
code.
Instead, an inexpensive dynamic disjointness check is performed by the Legion
runtime, and the safety of parallel execution follows from the soundness of
the type system.

### 2012-07-11 doc/popl2013/intro.tex
changes to intro and code example sections

In Legion data is organized in a hierarchy of {\em regions}
and subregions while computation is organized in a hierarchy of {\em
tasks} and subtasks operating on regions.  Regions and tasks interact
through a static system of {\em region permissions} specifying 
operations a task may perform on a region argument (read,
write, or reduce) and  {\em region coherence} annotations that
express what other tasks may do concurrently with
the region (exclusive, atomic, or simultaneous).  We prove the
soundness of the permissions and coherence system and use these two
results to prove a third result: if two {\em siblings} (tasks that
have the same immediate parent task) $t_1$ and $t_2$ are 
{\em independent} (have no ordering requirements for correctness), 
then the unique ancestors of $t_1$ and $t_2$ in the task hierarchy 
are also independent.  This theorem is the basis for correct distributed
task scheduling in a Legion implementation, which is crucial for
performance; a centralized scheduler would be a bottleneck
because of the huge latencies involved in communicating in the target
class of machines.

@@

Our own experience with writing high-performance applications in both 
Sequoia and the current industry standard mix of MPI, shared-memory
threads, and CUDA, has taught us that for many practical situations a
single, fairly static partitioning is insufficient.  In practice, the
best way to partition data is often a function of the data
itself---i.e., the partitions need to be dynamically computed.
Furthermore, it is useful to allow mutliple partitions simultaneously of
the same data, providing different views on to that data. Thus, in Legion the 
handling of regions and partitioning of regions is much more dynamic than in 
previous designs.  In particular,
\begin{itemize}
\item  logical regions are first-class values in Legion
and may be dynamically allocated and stored in data structures;

@@

\item  a logical region may be dynamically partitioned in multiple different ways.
\end{itemize}
One consequence of these decisions is that, through partitioning, a
given datum may belong to multiple different regions.  For example, if
a region $R$ is disjointly partitioned three distinct ways 
and each partition allocates every element of $R$ to a
subregion (partitions need not be total in Legion), then every element
of $R$ is included in three different subregions.  This is why we use
the term {\em logical} regions: language-level regions serve to name
sets of data but do not imply anything about physical layout or placement
in the memory hierarchy.  There
is a separate system of {\em physical regions} at runtime that hold
copies of the data in the language-level logical regions.  In fact,
the Legion implementation routinely maintains multiple distinct
physical regions, each with a copy of the data in the corresponding
logical region, for performance reasons \cite{Legion12}.

@@

type CircuitPiece<rl,rw,rn> = rr[rpw,rpn,rg]
                            < wire_list<rl,rpw,rpn,rg>@rl, node_list<rl,rpn>@rl >         
                            where rpn #$\le$# rn and rg #$\le$# rn and rpw #$\le$# rw and
                                  rpn * rg and rn * rw and rl * rn and rl * rw

@@

The important difference bewteen Legion and the more static approaches
is that Legion allows aliasing, but does so in a structured way that minimizes
runtime overheads.  The key observation is that
alias analysis is hard statically but very easy dynamically and is relatively cheap
when done at the granularity of regions instead of individual memory
locations.  Thus, Legion falls between fully static systems, such as
Sequoia and DPJ, and fully dynamic approaches such as transactional
memory.  Legion still has a significant static component; in
particular, the safety of permissions is checked statically and region
pointer dereferences are checked statically.  The dynamic alias tests
for independence are done at the granularity of regions and one check on a region
itself often serves to prove the independence of many subsequent region
uses.  In contrast, because transactional memory has no mechanism
for grouping heap data into larger collections, it must test the
overlap between two sets of data on each individual memory access, which is
significantly more expensive.

>>>>

In Legion data is organized in a hierarchy of {\em regions}
and subregions while computation is organized in a hierarchy of {\em
tasks} and subtasks operating on regions.  Regions and tasks interact
through a static system of {\em region permissions} specifying 
operations a task may perform on a region argument (read,
write, or reduce) and  {\em region coherence} annotations that
express what other tasks may do concurrently with
the region (exclusive, atomic, or simultaneous).  We prove the
soundness of the permissions and coherence system and use these two
results to prove a third result: if two {\em siblings} (tasks that
have the same immediate parent task) $t_1$ and $t_2$ are 
{\em non-interfering} (have no ordering requirements for correctness), 
then the any descendent of $t_1$ in the task hierarchy is non-interfering
with any descendent of $t_2$.  This theorem is the basis for correct distributed
task scheduling in a Legion implementation, which is crucial for
performance; a centralized scheduler would be a bottleneck
because of the huge latencies involved in communicating in the target
class of machines.

@@

Our own experience with writing high-performance applications in both 
Sequoia and the current industry standard mix of MPI, shared-memory
threads, and CUDA, has taught us that for many practical situations a
single, fairly static partitioning is insufficient.  In practice, the
best way to partition data is often a function of the data
itself---i.e., the partitions need to be dynamically computed.
Furthermore, it is useful to allow multiple partitions simultaneously of
the same data, providing different views on to that data. Thus, in Legion the 
handling of regions and partitioning of regions is much more dynamic than in 
previous designs.  In particular,
\begin{itemize}
\item  logical regions are first-class values in Legion
and may be dynamically allocated and stored in data structures;

@@

\item  a logical region may be dynamically partitioned in multiple different ways.
\end{itemize}
One consequence of these decisions is that, through partitioning, a
given datum may belong to multiple different regions.  For example, if
a region $R$ is disjointly partitioned three distinct ways 
and each partition allocates every element in $R$ to a
subregion (partitions need not be total in Legion), then every element
in $R$ is included in three different subregions.  This is why we use
the term {\em logical} regions: language-level regions serve to name
sets of data but do not imply anything about physical layout or placement
in the memory hierarchy.  There
is a separate system of {\em physical regions} at runtime that hold
copies of the data in the language-level logical regions.  In fact,
the Legion implementation routinely maintains multiple distinct
physical regions, each with a copy of the data in the corresponding
logical region, for performance reasons \cite{Legion12}.

@@

type CircuitPiece<rl,rw,rn> = rr[rpw,rpn,rg]
                            < wire_list<rl,rpw,rpn,rg>@rl, node_list<rl,rpn>@rl >         
                            where rpn #$\le$# rn and rg #$\le$# rn and rpw #$\le$# rw and
                                  rn * rw and rl * rn and rl * rw

@@

The important difference bewteen Legion and the more static approaches
is that Legion allows aliasing, but does so in a structured way that minimizes
runtime overheads.  The key observation is that
static alias analysis is hard, but dynamic analysis is very easy
and relatively cheap
when done at the granularity of regions instead of individual memory
locations.  Thus, Legion falls between fully static systems, such as
Sequoia and DPJ, and fully dynamic approaches such as transactional
memory.  Legion still has a significant static component; in
particular, the safety of permissions is checked statically and region
pointer dereferences are checked statically.  The dynamic alias tests
for independence are done at the granularity of regions and one check 
on a region
often serves to prove the independence of many subsequent uses of that region.
In contrast, because transactional memory has no mechanism
for grouping heap data into larger collections, it must test the
overlap between two sets of data on each individual memory access, which is
significantly more expensive.
